Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?

### 2022-12-20 supp/index.qmd
<edit index>

Elliptical bifaces are prevalent at Mississipian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississipian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178]. In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken. Elliptical bifaces find their most dramatic expression as Mississippian and Caddoan exaggerated, symbolic weaponry. Mississippian Ramey knives, dating from c. CE 1000 to 1300 are found associated with Cahokia political dynamics. The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation. Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology is intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

2.    The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001]. Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264]. Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
  a)   Elapid and viper species will traverse the smaller size class through normal developmental stages before reaching adulthood, where their morphology is known to differ within and among species. If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among the adults. Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2. However, the analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been intended to convey motion, where flake scars on one face represent muscle constriction on one edge and muscle dilation on the other, with an inverted pattern on the opposite face. Thus, in intentionally flipping between the two faces---which might have been a component of the serpent/snake dance---dancers could convey motion with a static implement. Additionally, the parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.

>>>>

Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], and are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178]. In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken. Elliptical bifaces find their most dramatic expression as Mississippian and Caddoan exaggerated, symbolic weaponry. Mississippian Ramey knives, dating from c. CE 1000 to 1300 are found associated with Cahokia political dynamics. The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation. Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

2.    The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001]. Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264]. Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
  a)   Elapid and viper species will progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species. If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among adults. Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2. However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, where the pattern is inverted on the opposite face. Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement. The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
